Black-swan events in animal populations

### 2016-11-10 text/ms_pnas.tex
Reorder supplementary figures
Edits while responding to reviewers
Also strip spaces

\newcommand{\tabstats}{{}{28}}
\newcommand{\tabcausessupp}{{2}}
\newcommand{\figpriors}{{1}}
\newcommand{\figsimnu}{{2}}
\newcommand{\figsimprob}{{3}}
\newcommand{\figalt}{{4}}
\newcommand{\figaltpriors}{{5}}
\newcommand{\figcorrelates}{{6}}
\newcommand{\figskewnu}{{7}}
\newcommand{\figskewprojections}{{8}}
\mathchardef\mhyphen="2D % math hyphen

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn@uw.edu}

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012} including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.

There are a number of caveats when considering the generality of our results.
The GPDD data represent a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.

@@

@@

One alternative approach would involve fitting
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
the GEV requires maxima per time block
or peaks beyond a chosen threshold.
Therefore, the GEV requires longer time series then typically available
for wild animal population abundance.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

For the Gompertz model,
we chose weakly informative priors for 
all parameters (Fig.~S\figpriors, Supporting Information).
For the degrees of freedom parameter $\nu$, 
we chose an exponential prior 
with a rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

@@

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).

We investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$; Fig.~S\figpriors, Supporting Information). Furthermore, we used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

>>>>

\newcommand{\tabstats}{{}{28}}
\newcommand{\tabcausessupp}{{2}}
\newcommand{\figalt}{{1}}
\newcommand{\figaltpriors}{{2}}
\newcommand{\figsimnu}{{3}}
\newcommand{\figsimprob}{{4}}
\newcommand{\figcorrelates}{{5}}
\newcommand{\figskewnu}{{6}}
\newcommand{\figskewprojections}{{7}}
\newcommand{\figpriors}{{8}}
\mathchardef\mhyphen="2D % math hyphen

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.}
\authordeclaration{The authors declare no conflict of interest.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn@uw.edu}

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012} including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction.
Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.
We expect that it is extrinsic forces that lead to the
prevalence of skew-$t$ distributed process noise:
there is no limit to how fast a population can decline due to a catastrophe,
but population increases are constrained
by the maximum biological rate of increase
possible for a given set of demographic traits
(e.g. population birth rate, mortality, age at maturity).

There are a number of caveats when considering the generality of our results.
The Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.

@@

TODO add assumptions here: assuming that productivity and density dependence are not time varying, this could be an important development for understanding the dynamics of longer time series

@@

One alternative approach would involve fitting
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
(e.g.\ sedimentation rates or wind speeds \cite{gaines1993, denny2009}),
the GEV requires maxima per time block.
Therefore, the GEV requires longer or higher frequency time series than typically available
for wild animal population abundance.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

For the Gompertz model,
we chose weakly informative priors for
all parameters (Fig.~S\figpriors, Supporting Information).
For the degrees of freedom parameter $\nu$,
we chose an exponential prior
with a rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\) since
below $2$ there can be issues with computational stability.
This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

@@

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence.
This model is equivalent to a random walk with drift
model on log abundance, and therefore does not assume
that the timeseries is stationary.
Third, observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).

We investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$) and alternative priors (Fig.~S\figpriors, Supporting Information).
Furthermore, we used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).
